---
title: "Research"
---

```{r packages, include=FALSE, message=FALSE, warning=FALSE}
library("scholar")
library("dplyr")
library("DT")
```

## Research Interests

I hold a PhD specialising in outbreak response modelling and decision-making. My work focuses on developing methods and open-source tools to evaluate the impact of pharmaceutical and non-pharmaceutical interventions on the spread of vaccine-preventable diseases.

**Key interests:**

- Compartmental modelling to quantify intervention impacts during epidemics.
- Real-time forecasting and nowcasting of infections and $R_t$ using Bayesian methods with tools like [EpiNow2](https://epiforecasts.io/EpiNow2/dev/), [epinowcast](https://package.epinowcast.org/dev/), [epidemia](https://imperialcollegelondon.github.io/epidemia/index.html), and [EpiEstim](https://mrc-ide.github.io/EpiEstim/).
- Building reproducible pipelines and open source packages for outbreak analytics and model implementation.
- Improving science research culture.

## Published work

You can find my published work on my [Google Scholar profile](https://scholar.google.co.za/citations?user=IxRpXp8AAAAJ&hl=en).

### Recent work

```{r pubs-table, echo=FALSE, message=FALSE, warning=FALSE}
# My scholar id
id <- "IxRpXp8AAAAJ"
# Get the profile
my_profile <- scholar::get_profile(id)
# Get publications
pubs <- scholar::get_publications(id)
# Sort by year
pubs <- pubs |> dplyr::arrange(desc(year))
# Get the first 5 publications
recent_pubs <- pubs |>
  dplyr::select(title, author, year, journal) |> 
  dplyr::slice_head(n = 5)
DT::datatable(recent_pubs)
```

### Top 5 most cited publications

```{r top-cited-pubs, echo=FALSE, message=FALSE, warning=FALSE}
# Get the top 5 most cited publications
# Sort by year
pubs <- pubs |> dplyr::arrange(desc(cites))
# Get the first 5 publications
most_cited <- pubs |>
  dplyr::select(title, author, year, journal, cites) |> 
  dplyr::slice_head(n = 5)
DT::datatable(most_cited)
```

## Ongoing research work

- [Evaluating the tradeoffs of using temporally aggregated data for epidemiological forecasts: a case study of COVID-19 in South Africa](https://github.com/tatendashoko/now-fore-cast).
- [A surveillance pipeline](https://github.com/SACEMA/severity-monitoring) to detect changes in reported COVID-19 severity (case fatality, case hospitalization, and hospital fatality).
- [Age-structured compartmental model](https://github.com/SACEMA/sa-polio-model) for estimating the timing of detection of polio-related acute flaccid paralysis in South Africa (generalizable to other settings).
- [A two-strain compartmental model](https://github.com/jamesmbaazam/two_strain_orv_model) 
for studying the impact of the timing of the emergence of variants on control strategies.

